Bayesian Parameter Inference for Nonlinear Stochastic Differential Equation Models

### 2015-08-19 TeXFiles/Manuscript_Albert_Ulzega.tex
incomplete new paper version (Simone)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent
% ( open parenthesis    ) close paren.
% - hyphen              = equals sign
% | vertical bar        ~ tilde
% @ at sign             _ underscore
% { open curly brace    } close curly
% [ open square         ] close square bracket
% + plus sign           ; semi-colon
% * asterisk            : colon
% < open angle bracket  > close angle
% , comma               . full stop
% ? question mark       / forward slash
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ
% abcdefghijklmnopqrstuvwxyz
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt,a4paper,final]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
\usepackage{iopams}
\usepackage{graphicx}
\usepackage[breaklinks=true,colorlinks=true,linkcolor=blue,urlcolor=blue,citecolor=blue]{hyperref}
%\usepackage{amsmath,amssymb,amsthm,amscd}
%\usepackage{natbib}

@@

If we denote the parameter vector $\bt=(\beta,\gamma)^T$ and assume a flat prior as described above, the posterior (\ref{Bayes}), as a function of $\bt$, is proportional to the likelihood function, which is expressed as a path-integral as follows
\begin{equation}\label{posterior_pathint}
  f_{\mbox{\small post}}(\bt | \vc y)
  \propto
  \int
  \exp\bigg[
    -\frac{1}{2}
    \sum_{s=1}^{n+1}
    \frac
    {(\ln(y_s/r(t_s))-{\beta q(t_s)})^2}
    {\sigma^2} -{{\mathcal S}}[{q},\dot q]
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,.
\end{equation}
On the right-hand side of the above equation, the first term describes the probability distribution of model outputs, given model parameters, rain inputs and a system realization $q(t_s)$. The second term is the probability of a system realization $q$ and the path-integral extends over all possible system realizations.

@@

In order to derive an efficient algorithm to draw parameter samples from (\ref{posterior_pathint}) we interpret it as the partition function of a 1D statistical mechanical system and simulate the dynamics of the latter, employing the so-called {\em Hamiltonian Monte Carlo} (HMC) algorithm \cite{duane_1987}.
The model parameters $\bt$ are interpreted as additional dynamical degrees of freedom coupling to the field $q(t)$.
Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian
\begin{equation}\label{Hamiltonian}
    \mathcal{H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})
    =
    K( p,{\pmb\pi}) + V( q,\bt)\,,
\end{equation}
where
\begin{equation}\label{K}
   K( p,{\pmb\pi})
   =
   \int_0^T \frac{ p^2(t)}{2m(t)}dt
   + \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,
\end{equation}
and $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
The posterior (\ref{posterior_pathint}) is expressed by the phase space path integral
\begin{equation}\label{phaseSpacePathInt}
    f_{\mbox{\small post}}(\bt | \vc y)
  \propto
  \int
  e^{-\mathcal {H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})}
  {\mathcal Dp}
   {\mathcal Dq}
   d{\pmb\pi}
  \,.
\end{equation}

The HMC method, which is a combination of the {\em Metropolis algorithm} \cite{metropolis_1953} and {\em molecular dynamics} methods \cite{alder_1959_MD, rahman_1964_MD}, iterates the following steps:
\begin{enumerate}
  \item
  Momenta are sampled from the Gaussian distribution defined by (\ref{K}).
  \item
  The system is integrated by means of a volume-preserving and time-reversible solution of a discretization of Hamilton's equations.
  \item
  The discretization error on the energy preservation due to the previous step is corrected by a Metropolis acceptance/rejection step.
\end{enumerate}
The last step is the standard Metropolis algorithm, while the first two steps allow us to make arbitrarily large jumps in phase space while maintaining an arbitrarily large acceptance rate.

In order to simulate the dynamics of Hamiltonian (\ref{Hamiltonian}), we need to discretize the path-integral (\ref{phaseSpacePathInt}).
Therefore, let us assume that the measurement time points $\left\{ y_s \right\}_{s=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
Each interval between two consecutive data points is further partitioned into $j$ bins, such that we have a total of $nj+1=N>>1$ discretization points.
Path-integral (\ref{phaseSpacePathInt}) is approximated by an ordinary integral, with the approximate path-measure
\begin{equation}
  \mathcal Dp\mathcal Dq
  \approx
  \prod_i dp_i dq_i
\end{equation}
and the discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$
\begin{eqnarray}
   K( p,{\pmb\pi}) \approx
   \sum_{i=1}^N
   \frac{ p_i^2}{2m_q}\Delta t
   +
   \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
   \\
   V(q,\bt) \approx \frac{\Delta t}{T} \sum_{i=2}^{N}
   \left\{ \frac{1}{2} T^2 \dot q_i^2 + \frac{1}{2}
     \left( \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i} \right)^2 -
    \frac{\beta^2}{2\gamma} e^{-\beta q_i} - T q_i\dot\rho_i \right\}  \nonumber
  \\
  +
  \frac{1}{\gamma}
  e^{-\beta q_N}
  +
  q_N \rho_{N}
  -
  \frac{1}{\gamma}
  e^{-\beta q_1}
  -
  q_1 \rho_{1}
  \nonumber
  \\
  +
  \sum_{s=1}^{n+1}
  \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta q_{(s-1)j+1}})^2}{2\sigma^2}
  \label{Vdisc}
   \,,
\end{eqnarray}
with
\begin{equation}
  \dot q_i = \frac{q_i-q_{i-1}}{\Delta t}\,,
  %\quad
%  \bar q_i = \frac{q_i+q_{i-1}}{2}\,,
\end{equation}
and
\begin{equation}\label{rhodisc}
\rho_i = \frac{T}{\beta} \frac{\ln(r(t_{i})/r(t_{i-1}))}{\Delta t}
+
\frac{(2+\gamma)\beta}{2\gamma}
\,,\quad
\dot\rho_i = \frac{\rho_i-\rho_{i-1}}{\Delta t}\,.
\end{equation}
Note that we've neglected terms of order $\mathcal O(N^{-1/2})$ in the action (\ref{Vdisc}).

The discretized Hamiltonian describes a {\em classical polymer chain} with harmonic bonds between neighbouring beads in an external field. The latter consists of two parts, a field that results from the dynamics of the original equations (\ref{sde}) and is felt by all the beads, and a field that results from the measurements and is felt by the measurement beads only.

The potential (\ref{Vdisc}) contains terms with different scaling in the potentially large numbers $N$ and $n$, which describe dynamics on different time-scales.
In particular, for large $N$, $V(q,\bt)$ is dominated by its harmonic part and a brute force numerical integration of Hamilton's equations in step (iii) of the HMC algorithm would require a very small discretization time-step to sufficiently resolve its dynamics.
An interesting approximative approach would be to employ a partial averaging of the fast Fourier modes as described in \cite{doll_1985_fourier}.
We choose an exact approach and employ a multiple time scale integration based on Trotter's formula \cite{tuckerman_1993}.

For this purpose it proves to be useful to introduce so-called \emph{staging} variables \cite{tuckerman_1992}, and diagonalize the harmonic part in between the measurement points.
Therefore, we rewrite the discretized harmonic part of the action as
\begin{eqnarray}
  \sum_{i=2}^{N}
  \frac{T}{2\Delta t}
  (q_i-q_{i-1})^2 \nonumber
  \\ =
  \frac{T}{2}
  \sum_{s=1}^{n}\left\{
    \frac{(q_{(s-1)j+1} - q_{sj+1})^2}{j\Delta t}
    +
    \sum_{k=2}^j
    \frac{k}{(k-1)\Delta t}
    (q_{(s-1)j+k}-q^*_{(s-1)j+k})^2
  \right\}\,,\nonumber
  \\
\end{eqnarray}
with
\begin{equation}
  q^*_{(s-1)j+k}
  =
  \frac{(k-1)q_{(s-1)j+k+1} + q_{(s-1)j+1} }{k}
  \,.
\end{equation}
We apply the coordinate transformation
\begin{equation}
  u_{sj+1} = q_{sj+1}\,,\quad
  u_{sj+k} = q_{sj+k} - q^*_{sj+k}\,,
\end{equation}
with its inverse given by
\begin{equation}
  q_{sj+1} = u_{sj+1}\,,\quad
  q_{sj+k} = \sum_{l=k}^{j+1}\frac{k-1}{l-1}u_{sj+l}
  +\frac{j-k+1}{j}u_{sj+1}\,,
\end{equation}
or, equivalently, by the recursive relation
\begin{equation}
  q_{sj+k} = u_{sj+k} + \frac{k-1}{k} q_{sj+k+1}+ \frac{1}{k}u_{sj+1} \,.
\end{equation}

@@

We use the Trotter formula according to \cite{tuckerman_1992} in order to design a reversible molecular dynamics integrator that takes these different time scales into account.
In order to design an appropriate partition of the Hamiltonian, we need to distinguish different regimes, such as
\begin{enumerate}
  \item[\it i.]
  $\mathcal H_N \sim \mathcal H_n >> \mathcal H_1$,
  \item[\it ii.]
  $\mathcal H_N >> \mathcal H_n \sim \mathcal H_1$,
  \item[\it iii.]
  $\mathcal H_N >> \mathcal H_n >> \mathcal H_1$.
\end{enumerate}
Here, we restrict ourselves to regime ($ii$), i.e., we assume that we do not have too many measurements and/or the measurement error is not too small.
The generalization of the method to the other schemes is straightforward.
In regime ($ii$) we simply separate the harmonic part of the action, for the staging beads, from the rest and write
\begin{equation}
  \mathcal{H}_{\mbox{\tiny HMC}}=\mathcal H_N + \mathcal H'\,.
\end{equation}
In order to design a reversible integrator, for the associated Hamiltonian equations, we define the Liouville operators
\begin{equation}
  iL_N=\{\cdot\,,\,\mathcal H_N\}\,,\quad
  iL'=\{\cdot\,,\,\mathcal H'\}\,,
\end{equation}
where $\{\cdot\,,\,\cdot\}$ denotes the Poisson brackets that are defined on functions on the phase space.
Trotter's formula \cite{trotter_1959} allows us to write the Hamiltonian propagator as
\begin{equation}
  e^{i(L_N+L')\tau}
  =
  (e^{iL_N(\Delta\tau/2)}e^{iL'\Delta\tau}e^{iL_N(\Delta\tau/2)})^P
  +
  \mathcal O(\tau^3/P^{2})\,,
\end{equation}
for $\tau =P\Delta \tau$.
In regime ($ii$) the outer propagator $\exp[iL_N(\Delta \tau/2)]$ describes a much faster dynamics than the inner one.
However, it is the dynamics of uncoupled harmonic oscillators, which we can readily solve.
Masses and frequencies of the oscillators are derived from (\ref{H_N}) as
\begin{equation}
  m=m_q/\Delta t\,,\quad
  \omega_k=\sqrt{\frac{Nk}{(k-1)m}}\,.
\end{equation}
The fast outer propagator is then given by the equations,
\begin{eqnarray}
  u_{(s-1)j+k}(\Delta\tau/2) \nonumber \\
  = u_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  +
  \frac{p_{(s-1)j+k}(0)}{m\omega_k}\sin(\omega_k\Delta\tau/2)\,,\\
  p_{(s-1)j+k}(\Delta\tau/2) \nonumber \\
  = p_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  -
  m\omega_k u_{(s-1)j+k}(0) \sin(\omega_k\Delta\tau/2)\,,
\end{eqnarray}
for $s=1,\dots,n$ and $k=2,\dots,j$.

For the inner, slow propagator, we employ the velocity Verlet algorithm \cite{swope_1982_verlet}.
For the boundary beads, it reads
\begin{eqnarray}
  u_{(s-1)j+1}(\Delta\tau) \nonumber \\
  = u_{(s-1)j+1}(0)
  +
  \frac{\Delta\tau}{m} p_{(s-1)j+1}(0)
  +
  \frac{\Delta \tau^2}{2m}
  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]\,,\\
  p_{(s-1)j+1}(\Delta\tau) \nonumber \\
  = p_{(s-1)j+1}(0)
  +
  \frac{\Delta\tau}{2}
  (
  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]
  +
  F_{(s-1)j+1}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
  )\,,
\end{eqnarray}
where $F_i[\vc u,{\pmb\theta}]$ denotes the partial derivative of $\mathcal H'[\vc u,{\pmb\theta}]$ w.r.t. $u_i$.
For the model parameters, analogous equations apply.
When it comes to the staging beads, however, only the momenta have to be updated, because the associated kinetic term is not part of $\mathcal H'$ but of $\mathcal H_N$.
Thus,
\begin{eqnarray}
  p_{(s-1)j+k}(\Delta\tau) \nonumber \\
  =
  p_{(s-1)j+k}(0)
  +
  \frac{\Delta\tau}{2}
  (
  F_{(s-1)j+k}[\vc u(0),{\pmb\theta}(0)]
  +
  F_{(s-1)j+k}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
  )
  \,.
\end{eqnarray}

\section{Implementation}

AD

parallelization

\section{Numerical Results}

The results shown in the figures refer to synthetic data obtained assuming a simple sinusoidal model for the rain input. The experimental data were then generated using the "true" parameter values $K=200$ and $\gamma = 0.2$. We considered $n=51$ data points, $j=10$ staging points, and therefore $N=501$ discretization points. The total time $T$ is $\approx 833$ [units of time], so that the discretization points are separated by a time interval $\Delta t \approx 1.6 $ [units of time]. Following the RESPA method described above, we used a long time step $\Delta \tau = 0.02$ [units of time] for the evolution of the slow (long-range) potential, further subdivided in 16 smaller steps for the evolution of the fast (short-range) potential.

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.4\textwidth]{Figs/FigRainData.png}
    \caption{\label{fig:rain_data_S}Rain input (dashed), system realization (solid line) and synthetic experimental data. The system output and the observed data were generated using $K=200$ and $\gamma = 0.2$.}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent
% ( open parenthesis    ) close paren.
% - hyphen              = equals sign
% | vertical bar        ~ tilde
% @ at sign             _ underscore
% { open curly brace    } close curly
% [ open square         ] close square bracket
% + plus sign           ; semi-colon
% * asterisk            : colon
% < open angle bracket  > close angle
% , comma               . full stop
% ? question mark       / forward slash
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ
% abcdefghijklmnopqrstuvwxyz
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt,a4paper,final]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
\usepackage{iopams}
\usepackage{graphicx}
\usepackage[breaklinks=true,colorlinks=true,linkcolor=blue,urlcolor=blue,citecolor=blue]{hyperref}
\usepackage{url}
\usepackage{caption}
%\usepackage{amsmath,amssymb,amsthm,amscd}
%\usepackage{natbib}

@@

If we denote the parameter vector $\bt=(\beta,\gamma)^T$ and assume a flat prior as described above, the posterior (\ref{Bayes}), as a function of $\bt$, is proportional to the likelihood function, which is expressed as a path-integral as follows
\begin{equation}\label{posterior_pathint}
  f_{\mbox{\small post}}(\bt | \vc y)
  \propto
  \int
  \exp\bigg[
    -\frac{1}{2}
    \sum_{s=1}^{n+1}
    \frac
    {(\ln(y_s/r(t_s))-{\beta q(t_s)})^2}
    {\sigma^2} -{{\mathcal S}}[{q},\dot q]
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,.
\end{equation}
On the right-hand side of the above equation, the first term describes the probability distribution of model outputs, given model parameters, rain inputs and a system realization $q(t_s)$. The second term is the probability of the given system realization $q$ and the path-integral extends over all possible system realizations. Instead of undertaking an often prohibitive numerical computation of such integral, we apply a Hamiltonian Monte Carlo method, outlined in the next section, to sample parameter vectors from a joint distribution of system realizations and model parameters given by an appropriate discretization of the action of the path-integral.

@@

In order to derive an efficient algorithm to draw parameter samples from (\ref{posterior_pathint}) we interpret it as the partition function of a 1D statistical mechanical system and simulate the dynamics of the latter, employing the so-called {\em Hamiltonian Monte Carlo} (HMC) algorithm \cite{duane_1987}.
The model parameters $\bt$ are interpreted as additional dynamical degrees of freedom coupling to the system variables $q(t)$.
Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian
\begin{equation}\label{Hamiltonian}
    \mathcal{H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})
    =
    K( p,{\pmb\pi}) + V( q,\bt)\,,
\end{equation}
where
\begin{equation}\label{K}
   K( p,{\pmb\pi})
   =
   \int_0^T \frac{ p^2(t)}{2m(t)}dt
   + \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,
\end{equation}
and $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
The posterior (\ref{posterior_pathint}) can then be expressed by the phase space path integral
\begin{equation}\label{phaseSpacePathInt}
    f_{\mbox{\small post}}(\bt | \vc y)
  \propto
  \int
  e^{-\mathcal {H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})}
  {\mathcal Dp}
   {\mathcal Dq}
   d{\pmb\pi}
  \,.
\end{equation}

The HMC method, which is a combination of the {\em Metropolis algorithm} \cite{metropolis_1953} and {\em molecular dynamics} methods \cite{alder_1959_MD, rahman_1964_MD}, iterates the following steps:
\begin{enumerate}
  \item
  Momenta $p$ and $\pi$ are sampled from the Gaussian distributions defined by eq.~(\ref{K}).
  \item
  The system is then allowed to evolve in $\left(q,\bt; p,{\pmb\pi}\right)$-phase space for an arbitrary time interval $\tau$ according to a volume-preserving and time-reversible solution of a discretization of Hamilton's equations.
  %The system is integrated by means of a volume-preserving and time-reversible %solution of a discretization of Hamilton's equations.
  \item
  The discretization error on the energy preservation due to the previous step is corrected by a Metropolis acceptance/rejection step.
\end{enumerate}
The last step is the standard Metropolis algorithm, while the first two steps allow us to make arbitrarily large jumps in phase space while maintaining an arbitrarily large acceptance rate. Each new phase space configuration corresponds to a well-defined combination of model parameters $\bt$. Thus, the simulated succession of configurations represent a Markov chain of sampled parameter vectors that naturally leads to the sought probability distribution for the system parameters.

In order to simulate the dynamics of the Hamiltonian (\ref{Hamiltonian}), we first need to discretize the path-integral (\ref{phaseSpacePathInt}).
Therefore, let us assume that the measurement time points $\left\{ y_s \right\}_{s=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
Each interval between two consecutive data points is further partitioned into $j$ bins, such that we have a total of $nj+1=N>>1$ discretization points.
The path-integral (\ref{phaseSpacePathInt}) is then approximated by an ordinary integral, with the approximate path-measure
\begin{equation}
  \mathcal Dp\mathcal Dq
  \approx
  \prod_i dp_i dq_i
\end{equation}
and the discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$
\begin{eqnarray}
   K( p,{\pmb\pi}) \approx
   \sum_{i=1}^N
   \frac{ p_i^2}{2m_q}\Delta t
   +
   \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
   \\
   V(q,\bt) \approx \frac{\Delta t}{T} \sum_{i=2}^{N}
   \left\{ \frac{1}{2} T^2 \dot q_i^2 + \frac{1}{2}
     \left( \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i} \right)^2 -
    \frac{\beta^2}{2\gamma} e^{-\beta q_i} - T q_i\dot\rho_i \right\}  \nonumber
  \\
  +
  \frac{1}{\gamma}
  e^{-\beta q_N}
  +
  q_N \rho_{N}
  -
  \frac{1}{\gamma}
  e^{-\beta q_1}
  -
  q_1 \rho_{1}
  \nonumber
  \\
  +
  \sum_{s=1}^{n+1}
  \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta q_{(s-1)j+1}})^2}{2\sigma^2}
  \label{Vdisc}
   \,,
\end{eqnarray}
with
\begin{equation}
  \dot q_i = \frac{q_i-q_{i-1}}{\Delta t}\,,
  %\quad
%  \bar q_i = \frac{q_i+q_{i-1}}{2}\,,
\end{equation}
and
\begin{equation}\label{rhodisc}
\rho_i = \frac{T}{\beta} \frac{\ln(r(t_{i})/r(t_{i-1}))}{\Delta t}
+
\frac{(2+\gamma)\beta}{2\gamma}
\,,\quad
\dot\rho_i = \frac{\rho_i-\rho_{i-1}}{\Delta t}\,.
\end{equation}
Note that we've neglected terms of order $\mathcal O(N^{-1/2})$ in the action (\ref{Vdisc}).

The discretized Hamiltonian describes a {\em classical polymer chain} of $N$ beads with harmonic bonds between neighbouring beads in an external field. The latter consists of two parts, a field that results from the dynamics of the original equations (\ref{sde}) and is felt by all the beads, and a field that results from the measurements and is felt by the measurement beads only. The masses $m_q$ and $m_{\alpha}$ are tunable parameters of the algorithm. A good strategy consists in assigning larger weights to the beads associated with measurement points than to the intermediate beads. The dynamics of the system will be therefore mostly dependent on the lighter intermediate particles, while the heavy measurement beads will be constrained in the vicinity of the measured data points. We thus reduce the original Bayesian inference problem to simulating the dynamics of a linear polymer, as sketched in Fig.~(\ref{fig:polymer}). In this context the physical time (horizontal axis) is interpreted as a spatial dimension while a fictitious simulation time is introduced (vertical axis). Each \emph{state} of this fictitious molecule corresponds to a well-defined configuration in the original phase space, characterized by a set of system variables $\left\{q_i\right\}_{i=1,\dots,N}$ and a parameter vector $\bt$.

\begin{figure}[htb!]
    \centering
    \captionsetup{justification=justified}
    \includegraphics[width=0.7\textwidth]{Figs/FigPolymer.png}
    \caption{Simulated dynamics of a polymer chain with $n+1=11$ data points (large filled circles) and $j-1=9$ intermediate beads (small empty circles). All other parameters are discussed in Section~\ref{numerical_results}. Top: initial state. The intermediate beads represent a linear interpolation to the data points. Bottom: polymer configuration after 1000 iterations of the propagation algorithm. The dotted line represents the initial configuration. It is evident that the new configuration is mostly determined by the dynamics of the low-mass intermediate beads, while the heavy-mass data points barely move from their initial positions.}
    \label{fig:polymer}
\end{figure}

The potential (\ref{Vdisc}) contains terms with different scaling in the potentially large numbers $N$ and $n$, which describe dynamics on different time-scales.
In particular, for large $N$, $V(q,\bt)$ is dominated by its harmonic part and a brute force numerical integration of Hamilton's equations in step (ii) of the HMC algorithm would require a very small discretization time-step to sufficiently resolve its dynamics.
An interesting approximative approach would be to employ a partial averaging of the fast Fourier modes as described in \cite{doll_1985_fourier}.
We choose an exact approach and employ a multiple time scale integration based on Trotter's formula \cite{tuckerman_1993}.

For this purpose it proves to be useful to introduce so-called \emph{staging} variables \cite{tuckerman_1992}, and diagonalize the harmonic part in between the measurement points.
Therefore, we rewrite the discretized harmonic part of the action as
\begin{eqnarray}
  \sum_{i=2}^{N}
  \frac{T}{2\Delta t}
  (q_i-q_{i-1})^2 \nonumber
  \\ =
  \frac{T}{2}
  \sum_{s=1}^{n}\left\{
    \frac{(q_{(s-1)j+1} - q_{sj+1})^2}{j\Delta t}
    +
    \sum_{k=2}^j
    \frac{k}{(k-1)\Delta t}
    (q_{(s-1)j+k}-q^*_{(s-1)j+k})^2
  \right\}\,,\nonumber
  \\
\end{eqnarray}
with
\begin{equation}
  q^*_{(s-1)j+k}
  =
  \frac{(k-1)q_{(s-1)j+k+1} + q_{(s-1)j+1} }{k}
  \,.
\end{equation}
We apply the coordinate transformations
\begin{equation}
  u_{sj+1} = q_{sj+1}\,, \qquad\qquad s=0,\dots,n\, ,
\end{equation}
for the boundary beads, corresponding to the original measurement points, and
\begin{equation}
  u_{sj+k} = q_{sj+k} - q^*_{sj+k}\,, \quad s=0,\dots,n-1\, , \quad k=2,\dots,j\, ,
\end{equation}
for the intermediate staging beads.
The inverse transformations are given by
\begin{eqnarray}
  q_{sj+1} = u_{sj+1}\,,
  \\
  q_{sj+k} = \sum_{l=k}^{j+1}\frac{k-1}{l-1}u_{sj+l}
  +\frac{j-k+1}{j}u_{sj+1}\,.
\end{eqnarray}
The latter can be equivalently expressed by the recursive relation
\begin{equation}
  q_{sj+k} = u_{sj+k} + \frac{k-1}{k} q_{sj+k+1}+ \frac{1}{k}u_{sj+1} \,.
\end{equation}

@@

We use the Trotter formula according to \cite{tuckerman_1992} in order to design a reversible molecular dynamics integrator that takes these different time scales into account.
In order to design an appropriate partition of the Hamiltonian, we need to distinguish different regimes, such as
\begin{enumerate}
  \item[\it i.]
  $\mathcal H_N \sim \mathcal H_n >> \mathcal H_1$,
  \item[\it ii.]
  $\mathcal H_N >> \mathcal H_n \sim \mathcal H_1$,
  \item[\it iii.]
  $\mathcal H_N >> \mathcal H_n >> \mathcal H_1$.
\end{enumerate}
Here, we restrict ourselves to regime ($ii$), i.e., we assume that the number of measurements $n$ is small and/or the measurement error $\sigma$ is large.
The generalization of the method to the other schemes is straightforward.
In regime ($ii$) we simply separate the harmonic part of the action, for the staging beads, from the rest and write
\begin{equation}
  \mathcal{H}_{\mbox{\tiny HMC}}=\mathcal H_N + \mathcal H'\,.
\end{equation}
In order to design a reversible integrator for the associated Hamilton equations, we define the Liouville operators
\begin{equation}
  iL_N=\{\cdot\,,\,\mathcal H_N\}\,,\quad
  iL'=\{\cdot\,,\,\mathcal H'\}\,,
\end{equation}
where $\{\cdot\,,\,\cdot\}$ denotes the Poisson brackets that are defined on functions on the phase space.
Trotter's formula \cite{trotter_1959} allows us to write the Hamiltonian propagator as
\begin{equation}\label{propagator}
  e^{i(L_N+L')\tau}
  =
  (e^{iL_N(\Delta\tau/2)}e^{iL'\Delta\tau}e^{iL_N(\Delta\tau/2)})^P
  +
  \mathcal O(\tau^3/P^{2})\,,
\end{equation}
for $\tau =P\Delta \tau$.
In regime ($ii$) the outer propagator $\exp[iL_N(\Delta \tau/2)]$ describes much faster dynamics than the inner one. However, it is the dynamics of uncoupled harmonic oscillators, which we can readily solve.
Masses and frequencies of the oscillators are derived from (\ref{H_N}) as
\begin{equation}
  m=m_q/\Delta t\,,\quad
  \omega_k=\sqrt{\frac{Nk}{(k-1)m}}\,.
\end{equation}
The fast outer propagator is then given by the equations,
\begin{eqnarray}
  u_{(s-1)j+k}(\Delta\tau/2) \nonumber \\
  = u_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  +
  \frac{p_{(s-1)j+k}(0)}{m\omega_k}\sin(\omega_k\Delta\tau/2)\,,\\
  p_{(s-1)j+k}(\Delta\tau/2) \nonumber \\
  = p_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  -
  m\omega_k u_{(s-1)j+k}(0) \sin(\omega_k\Delta\tau/2)\,,
\end{eqnarray}
for $s=1,\dots,n$ and $k=2,\dots,j$.

For the inner, slow propagator, we employ the velocity Verlet algorithm \cite{swope_1982_verlet}.
For the boundary beads, it reads
\begin{eqnarray}
  u_{(s-1)j+1}(\Delta\tau) \nonumber \\
  = u_{(s-1)j+1}(0)
  +
  \frac{\Delta\tau}{m} p_{(s-1)j+1}(0)
  +
  \frac{\Delta \tau^2}{2m}
  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]\,,\\
  p_{(s-1)j+1}(\Delta\tau) \nonumber \\
  = p_{(s-1)j+1}(0)
  +
  \frac{\Delta\tau}{2}
  (
  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]
  +
  F_{(s-1)j+1}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
  )\,,
\end{eqnarray}
with $s=1,\dots,n+1$ and where $F_i[\vc u,{\pmb\theta}]$ denotes the partial derivative of $\mathcal H'[\vc u,{\pmb\theta}]$ w.r.t. $u_i$.
For the model parameters $\bt$ and their momenta ${\pmb\pi}$, analogous equations apply.
When it comes to the staging beads, however, only the momenta have to be updated, because the associated kinetic term is not part of $\mathcal H'$ but of $\mathcal H_N$.
Thus,
\begin{eqnarray}
  p_{(s-1)j+k}(\Delta\tau) \nonumber \\
  =
  p_{(s-1)j+k}(0)
  +
  \frac{\Delta\tau}{2}
  (
  F_{(s-1)j+k}[\vc u(0),{\pmb\theta}(0)]
  +
  F_{(s-1)j+k}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
  )
  \,,
\end{eqnarray}
with $s=1,\dots,n$ and $k=2,\dots,j$. The above propagators are applied sequentially $P$ times to calculate the system evolution over the time $\tau$.

\section{Numerical Results}\label{numerical_results}

The algorithm described above was implemented using the free and open-source programming language \emph{Julia} (version 0.3.9)~\cite{julia_web}. One of the attractive features of the language is its (fast growing) library of over 600 packages for numerical and scientific applications.
In our case, we made extensive use of the \emph{ReverseDiffSource} package, which provides a powerful tool for reverse-mode automated differentiation (AD) of arbitrarily complex user-defined expressions. Our algorithm benefits greatly from the use of AD. Indeed, it gives us the possibility to modify the model of eq.~(\ref{sde}) and therefore the action~(\ref{action}) leaving the implementation of the algorithm unaltered. This clearly gives our program significant flexibility making it suitable for a much broader range of applications than the simple exemplary SDE model described here.
